Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-10 text/supp.tex
Edit edit edit

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an exponential rate parameter of $0.1$. We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined. This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be informative about the precise value of $\nu$. In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper (e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy tailed.

We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential scale reduction factor --- a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged. These chain lengths may seem low to those familiar with software such as WinBUGS or JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.

@@

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is known to affect the ability to detect extreme events \citep{ward2007}. In our main analysis, we fit a model that ignored observation error. One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error. However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model \citep{knape2008}. Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter --- the shape of the process deviation tails --- potentially making identifiability and computational issues even greater. Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), and $\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set $\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

We separated our simulation into these two components to avoid confounding two issues. (1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution. In that case, no model, no matter how perfect, will be able to detect the shape of the tails. (2) Complex models may return biased parameter estimates if there are conceptual, computational, or coding errors. Our first simulation tested the first issue and our second simulation tested the latter. In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events --- especially given the length of the time series in the GPDD.

@@

Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (TODO FIG).
The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness. Fitting a model with correctly specified observation error did not make substantial improvements to model bias.

@@

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well. Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed. The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, upper panels). The model would have correctly categorized all cases where the process noise was not heavy tailed (indicated as ``$\nu =$ infinity'' in Fig.~\ref{fig:sim-prob}) and all cases where $\nu = 3$. With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all cases where $\nu = 3$. Allowing for observation error made little improvement to the detection of heavy tails (Fig.~\ref{fig:sim-prob}, lower-left vs.\ lower-right panel). Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.

@@

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}b), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor. (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
\caption{
  (fishes, crustaceans, gastropods, sharks). All filtered time series used in our analysis. The abundances are shown on a log10 vertical axis. Throughout this figure, red dots indicate values that were interpolated and red circles indicate values that were recorded as zero but were set to the next lowest observed abundance. Numbers before each species name are the GPDD main identification numbers.
}
\label{fig:all-ts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/check-sim-box.pdf}
\caption{
Simulation testing the Gompertz estimation model when the process deviation
draws were chosen so that $\nu$ could be estimated close to the true value
outside the full population model (``effective $\nu$'' within a CV of 0.2 of
specified $\nu$).
Plots show the probability that $\nu < 10$ (i.e.\ the approximate probability
of heavy tails) across three population $\nu$ values
(3, 5, and normal) and different scenarios panels: (1) 100 time steps and no
observation error, (2) 50 time steps and no observation error, (3) 50 time
steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
ignored, and (4) 50 time steps with observation error in which the quantity
of observation error was assumed known.
Within
each scenario the dots represent stochastic draws from the true population
distributions combined with model fits.
}
\label{fig:sim-prob}
\end{center}
\end{figure}

>>>>

Our prior on $\nu$ was based on \citet{fernandez1998}. They chose an exponential rate parameter of $0.1$. We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined. This prior gives only a $7.7$\% probability that $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity---above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be informative about the precise value of $\nu$. In the scenario where the data are uninformative about heavy tails (e.g.~Fig.~\ref{fig:didactic}e,~h), the posterior will approximately match the prior (prior median $= 71$, mean $= 102$) and the metrics used in our paper (e.g.~Pr$(\nu < 10) > 0.5$) are unlikely to flag the population as heavy tailed.

We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. We began with four chains and $2000$ iterations, discarding the first $1000$ as warm up (i.e.~4000 total samples). If $\hat{R}$ (the potential scale reduction factor---a measure of chain convergence) was greater than $1.05$ for any parameter or the minimum effective sample size, $n_\mathrm{eff}$, (a measure of the effective number of uncorrelated samples) for any parameter was less than $200$, we doubled both the total iterations and warm up period and sampled from the model again. These thresholds are in excess of the minimums recommended by \citet{gelman2006a} of $\hat{R} < 1.1$ and effective sample size $> 100$ for reliable point estimates and confidence intervals. In the majority of cases our minimum thresholds were greatly exceeded. We continued this procedure up to $8000$ iterations ($16000$ total samples) by which all chains were deemed to have sufficiently converged. These chain lengths may seem low to those familiar with software such as WinBUGS or JAGS, but the No-U-Turn Hamiltonian Markov chain Monte Carlo Sampler in Stan generally requires far fewer iterations to obtain equivalent effective sample sizes \citep{stan-manual2014}.

@@

Observation error can bias parameter estimates \citep[e.g.][]{knape2012} and is known to affect the ability to detect extreme events \citep{ward2007}. In our main analysis, we fit a model that ignored observation error. One way to account for observation error would be to fit a full state-space model that simultaneously estimates the magnitude of process noise and observation error. However, simultaneously estimating observation and process noise is a challenging problem (e.g.\ because the observation and process noise parameters tend to negatively covary in model fitting) and is known to sometimes result in identifiability issues with the Gompertz population model \citep{knape2008}. Furthermore, our model was applied to hundreds of time series, often of short length (as few as 20 time steps) and our model estimates an additional parameter---the shape of the process deviation tails---potentially making identifiability and computational issues even greater. Therefore, we considered a version of the base Gompertz model that allowed for a fixed level of observation error:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align*}
where $U$ represents the unobserved state vector, $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), and $\sigma_\mathrm{proc}$ represent the process noise scale parameter. We set $\sigma_\mathrm{obs}$ to $0.2$, which represents the upper limit of values often used in simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

@@

We separated our simulation into these two components to avoid confounding two issues. (1) With smaller sample sizes, there may not be a stochastic draw from the tails of a distribution. In that case, no model, no matter how perfect, will be able to detect the shape of the tails. (2) Complex models may return biased parameter estimates if there are conceptual, computational, or coding errors. Our first simulation tested the first issue and our second simulation tested the latter. In general, our simulations show that, if anything, our model under predicts the magnitude and probability of heavy tailed events---especially given the length of the time series in the GPDD.

@@

Our results indicate that the Gompertz model can recapture the true value of $\nu$ when the process noise was chosen so that appropriate tail events were present (Fig.~\ref{fig:sim-prob} upper panels). The addition of observation error caused the model to tend to underestimate the degree of heavy-tailedness. Fitting a model with correctly specified observation error did not make substantial improvements to model bias (Fig.~\ref{fig:sim-prob}).

@@

When converting the posterior distributions of $\nu$ into Pr($\nu < 10$), the models distinguished heavy and not-heavy tails reasonably well (Fig.~\ref{fig:sim-prob} lower panels). Without observation error, and using a probability of $0.5$ as a threshold, the model correctly classified all simulated systems with normally distributed process noise as not heavy tailed. The model would have miscategorized only one of $40$ simulations at $\nu = 5$ across a sample size of $50$ or $100$ data points (Fig.~\ref{fig:sim-prob}, scenarios 1 and 2 in lower row, second panel from left). The model would have correctly categorized all cases where the process noise was not heavy tailed (Fig.~\ref{fig:sim-prob} bottom-right panel) and all cases where $\nu = 3$ and there was not observation error. With $0.2$ standard deviations of observation error, the model still categorized \obsErrorNuFivePerc\% of cases as heavy tailed when $\nu = 5$ and all but one case when $\nu = 3$. Allowing for observation error made little improvement to the detection of heavy tails. Therefore, we chose to focus on the simpler model without observation error in the main text, particularly given that the true magnitude of observation error was unknown in the empirical data.

@@

To derive taxonomic-order-level estimates of the probability of heavy tails accounting for time-series length (Fig.~\ref{fig:posteriors}a), we fit a separate multilevel model with the same structure but with only $\log$ time-series length as a predictor. (In this case, we did not want to control for intrinsic population characteristics such as density dependence.) Since our predictors were centered by subtracting their mean value, we obtained order-level estimates of the probability of heavy tails at mean log time-series length by adding the posteriors for $\alpha$, $\alpha^\mathrm{class}_j$, and $\alpha^\mathrm{order}_k$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-fishes-others.pdf}
\caption{
  (fishes, crustaceans, gastropods, sharks). All filtered time series used in our analysis. The abundances are shown on a log10 vertical axis. Throughout this figure, red dots indicate values that were interpolated and blue dots indicate values that were recorded as zero but were set to the next lowest observed abundance. Numbers before each species name are the GPDD Main ID numbers.
}
\label{fig:all-ts}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-median-dist.pdf}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-p10.pdf}
\caption{
Simulation testing the Gompertz estimation model when the process deviation
draws were chosen so that $\nu$ could be estimated close to the true value
outside the full population model (``effective $\nu$'' within a CV of 0.2 of
specified $\nu$).
Upper panels show the distribution of median $\widehat{\nu}$ across 20 simulation runs.
Lower panels show the distribution of Pr($\nu < 10$) across 20 simulation runs.
We ran the simulations across three population (``true'') $\nu$ values
(3, 5, and $1\cdot 10^9$, i.e.\ approximately normal) and four scenarios: (1) 100 time steps and no
observation error, (2) 50 time steps and no observation error, (3) 50 time
steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but
ignored, and (4) 50 time steps with observation error in which the quantity
of observation error was assumed known.
Within each scenario the dots represent stochastic draws from the true population
distributions combined with model fits.
}
\label{fig:sim-prob}
\end{center}
\end{figure}
